Inverse conformational selection in lipid-protein binding

### 2021-07-05 manuscriptPGPE.tex

\begin{abstract}
% insert abstract here
  Interest in lipid interactions with proteins and other biomolecules is emerging not only in
  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
  are commonly used, for example, as carriers in mRNA vaccines.
  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
  interact with approaching substances, such as proteins, drugs, RNA or viruses.
  %In the physiologically relevant liquid state,
  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
  Because conformational ensembles of lipids in physiologically relevant liquid state 
  %However, the distributions of conformations in such ensembles and their lipid-type dependence
  have not yet been experimentally determined, a crucial piece has been missing from the understanding
  on how they interact with other biomolecules: 
  It is not clear if lipid molecules exchange between a few rigid structures,
  or fluctuate freely across a practically continuous spectrum of  conformations.
  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions. We find that all lipid headgroup types sample a wide range of
  overlapping conformations % a wide range of conformations
  in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
  in probability distributions of conformations. %, but all four lipid types can access .
  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB),
  this suggests that all four key lipid types can bind to proteins in a wide range of conformations,
  which are not limited by the headgroup chemistry.
  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
  %lipid-mediated signaling.
  Our results propose that lipids are able to select a suitable conformation
  from the wide range available to them to fit the various binding sites in proteins.
  The proposed {\it Inverse Conformational Selection Model} will extend also to lipid binding
  to other targets than proteins, such as drugs, RNA and viruses.

@@

The heavy atom dihedral angle distributions calculated from these conformations 
(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit a wide
range of conformations, independent of the headgroup type.
As for bulk lipid bilayers (Figs.~\ref{structures}C, \ref{DIHdists}), only {\it cis} conformations
of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals are almost
completely absent in all lipids, and significant differences between
different headgroups are not observed.
Structures deviating from lipid crystals 
have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
However, we observe large deviations from the lipid crystal structures also in conformational ensembles
that reproduce the NMR data in the liquid lamellar phase, indicating that such deviations
are realistic also in the protein-bound states.

@@

Our results suggest that lipid binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
%We suggest that the key to understanding selective binding of certain lipid types to proteins is in intermolecular lipid--protein interactions, rather than in conformational restrictions inherent to the lipids. The broad conformational ensembles we see in bulk bilayers suggest that the structures observed in pure lipid crystals occur seldomly in liquid state. Consequently, in protein binding the entropic cost on the lipid may be significant.
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.

>>>>

\begin{abstract}
% insert abstract here
  Interest in lipid interactions with proteins and other biomolecules is emerging not only in
  fundamental biochemistry, but also in the field of nanobiotechnology where lipid nanoparticles
  are commonly used, for example, as carriers in mRNA vaccines.
  The outward facing components of lipid bilayer, the lipid headgroups, regulate how cellular membranes or lipid nanoparticles
  interact with approaching substances, such as proteins, drugs, RNA or viruses.
  %In the physiologically relevant liquid state,
  %lipid molecules are not fixed to a single structure but sample an ensemble of conformations with Boltzmann-weighted probabilities.
  Because conformational ensembles of lipid headgroups in physiologically relevant liquid state 
  %However, the distributions of conformations in such ensembles and their lipid-type dependence
  have not yet been experimentally determined, a crucial piece has been missing from the understanding
  on how they interact with other biomolecules: 
  It is not clear if lipid molecules exchange between a few rigid structures,
  or fluctuate freely across a practically continuous spectrum of  conformations.
  Here, we combine solid-state NMR experiments and molecular dynamics simulations from the NMRlipids Project to resolve the conformational ensembles of headgroups of four key lipid types in various biologically relevant conditions. We find that all lipid headgroup types sample a wide range of
  overlapping conformations % a wide range of conformations
  in both neutral and charged cellular membranes, and that differences in the headgroup chemistry manifests only
  in probability distributions of conformations. %, but all four lipid types can access .
  Together with the analysis of 894 protein-bound lipid structures from the Protein Data Bank (PDB),
  this suggests that all four key lipid types can bind to proteins in a wide range of conformations,
  which are not limited by the headgroup chemistry.
  %Therefore, the selective adsorption of peripheral proteins to membranes and the binding of lipids to transmembrane proteins are likely to be regulated by specific protein--lipid interactions, not by any inherent conformational restrictions of the lipids.
  %Our results pave the way to comprehensive understanding of lipid--protein interactions in biomedical applications, in particular in
  %lipid-mediated signaling.
  Our results propose that lipids are able to select a suitable headgroup conformation
  from the wide range available to them to fit the various binding sites in proteins.
  The proposed {\it Inverse Conformational Selection Model} will extend also to lipid binding
  to other targets than proteins, such as drugs, RNA and viruses.

@@

The heavy atom dihedral angle distributions calculated from these conformations 
(Fig.~\ref{dihedralsFROMpdb} A) reveal that the protein-bound lipids indeed exhibit a wide
range of conformations, independent of the headgroup type.
As for bulk lipid bilayers (Figs.~\ref{structures}C, \ref{DIHdists}), only {\it cis} conformations
of P-O$_\alpha$-C$_\alpha$-C$_\beta$ and g$_2$-g$_3$-O$_{g_3}$-P dihedrals are almost
completely absent in all lipids, and significant differences between
different headgroups are not observed.
Protein-bound lipid structures deviating from protein-free lipid crystal structures 
have been previously proposed to indicate inaccuracies in lipid structures in PDB \cite{marsh13b,pezeshkian18}. 
However, we observe large deviations from the protein-free lipid crystal structures also in conformational ensembles
that reproduce the NMR data in the liquid lamellar phase, indicating that such deviations
are realistic also in the protein-bound states.

@@

Our results suggest that lipid headgroup binding to proteins can be described by the {\it inverse conformational selection model}
where unbound lipids sample a wide conformational ensemble offering the large selection of conformations
that can bind to various binding sites in substrates, such as protein, drug, RNA, or virus, that remain conformationally fixed.
In this model, the binding energy is dominated by
intermolecular interactions and changes in lipid conformational entropy, while 
conformational restrictions inherent to the lipids play a minor role.
Here we have focused only on lipid headgroups and potential applications of the proposed model on interactions between hydrophobic acyl chains and proteins are left for future studies. 
Applications of the proposed model in molecular cell biology include the understanding of lipid-mediated cell signaling
and functions of lipid regulated membrane proteins in general. 
In bionanotechnology, our model facilitates, for example,
the design of phoshopolipid-specific antibodies and lipid nanoparticle carriers with minimal sides effects for mRNA vaccines.
Furthermore, our work has demonstrated the power of open access MD simulation data
to complement the PDB data in elucidating behavior of complex systems of disordered biomolecules.
